Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-08-15 latex/main.tex

@@

cell, press, LaTeX, template

@@

No subheadings, please. Use of abbreviations should be kept to a minimum, and nonstandard abbreviations should be defined when first used in the text.

@@

The results may be divided with subheadings. In our view, good subheadings convey information about the findings, so we encourage you to be specific. For example, "Factor X requires factor Y to function in process Z" is better than "Analysis of factors X and Y by approach Q." We recommend that you use similar language in your figure titles for clarity and structural harmony. Please do not use numbered subheadings. For some formats, ``results and discussion" may be combined under a single heading.

\subsection*{Figures}

For initial submissions, figures and their titles and legends can be run-in with the text. For final submissions, the image files must be uploaded as separate files, one complete file per figure. The figure titles and legends should be placed toward the end of the main text, after the declaration of interests. Our typesetters will place the figures and their titles and legends appropriately based on your in-text citations.

\subsection*{Citations}

To cite/link to display items (figures and tables) and/or sections of the manuscript, simply write, for example, ``(Figure 1)" or ``see discussion." Our team will do the rest. Please note that we do not number sections or subsections.

To cite references, you may use the cite command, e.g., ``Recent articles in \textit{Matter} and \textit{Cell} \cite{wang2020conformational,king2002,cates1984statics}  have shown ..." or  ``Many interesting discoveries have been reported, \cite{radzihovsky1995new,kapoor2022leakage,hantke2018my,alcock2006} which ..."

\subsection*{Equations}

Simple formulae should appear in line with the text whenever possible. You can write inline math by enclosing it between \verb|\(| and \verb|\)|, as in this example: \(x^2 + y^2 = z^2\). You can also enclose it between dollar signs (\texttt{\$}), as in this example: $E=mc^2$.

Larger, more complex formulae may appear on a new line, either by enclosing them between \verb|\[| and \verb|\]| or by using the \verb|displaymath| environment:

\[ x^n + y^n = z^n \]

\begin{displaymath}
\sqrt{x^2+1}
\end{displaymath}

If any equations or formulae need to be referred to or cited again later in the text, use the \verb|equation| environment to number them. Later, you can cite these as "Equation 1," "Equation 2," etc.

\begin{equation}
f(x) = \sum_{i=0}^{n} \frac{a_i}{1+x}
\end{equation}

@@

The discussion should explain the significance of the results and place them into a broader context. Subheadings are permitted.

@@

A "limitations" or "limitations of the study" subsection in the discussion is encouraged and may be required for some journals and some article formats.

>>>>

\graphicspath{ {../figures/} }

@@

ancestral recombination graphs, recombination, SARS-CoV-2

@@

%%% No subheadings, please. Use of abbreviations should be kept to a minimum, 
%%% and nonstandard abbreviations should be defined when first used in the text.

The genomic revolution has brought a paradigm shift in how we study the evolution and spread of pathogens, 
allowing real-time tracking of pathogens over time and space. 
During the COVID-19 pandemic, thousands of genome sequences of SARS-CoV-2 samples were generated daily, 
quickly overwhelming typical phylogenetic approaches. 
Although real-time phylogenetic methods now exist to cope with this flood of genomic data 
(e.g., UShER\cite{Turakhia2021,DeMaio2023}), we lack methods to build phylogenies 
that can integrate recombination in the face of ongoing sampling of infected cases.

UShER is a maximum parsimony method developed during the pandemic to infer phylogenies 
from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}. 
UShER has generated phylogenies updated daily \cite{McBroome2021} 
and proved to be a powerful resource that successfully enabled the world 
to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}). 
These phylogenies are used to identify and monitor 
potentially epidemiologically important “Pango” lineages \cite{Rambaut2020,McBroome2024}, 
which might have acquired features to become Variants of Concern (VOC), 
for example, increased transmissibility, disease severity, and immune escape.
Beyond supporting real-time genomic epidemiology, 
UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2 
(e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}). 
UShER uses a tree to represent evolutionary change via mutation, assuming no recombination \cite{Turakhia2021}. 
Given extensive evidence for recombination in SARS-CoV-2 \cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2022}, 
the tree-like representation of UShER is no longer tenable.

Ancestral Recombination Graphs (ARGs) are reticulate genealogies that incorporate both mutation and recombination events \cite{Wong2024}, 
potentially allowing more comprehensive evolutionary representations of SARS-CoV-2 and other viruses undergoing recombination. 
Until recently, ARGs lacked software support and sufficiently scalable inference methods to be of practical use. 
Recent advances enable ARG inference for tens to hundreds of thousands of human genomes \cite{Kelleher2019,Speidel2019,Zhang2023}, 
with high levels of recombination. 
These advances now enable evolutionary genealogies to be built quickly and at scale, 
and there is a surge of excitement in the inference and applications of ARGs \cite{Brandt2024,Lewanski2024,Wong2024,Nielsen2025}.

Previous ARG methods are retrospective \cite{Rasmussen2014,Kelleher2019,Speidel2019,Ignatieva2022,Zhang2023}, 
reconstructing the past from a sample of genome sequences drawn at one moment in time, 
but scientists are increasingly tracking evolutionary change as it happens, accumulating data forward in time. 
For example, sequences from daily samples of SARS-CoV-2 were made globally available on a daily basis, 
leading to over 17 million sequences in the GISAID database \cite{Shu2017}.

Here, we present 
(1) sc2ts, a new method to infer ARGs in real time by building genealogies forward in time, and 
(2) an ARG for SARS-CoV-2 with ~2.48 million genome sequences. We used the “Viridian” dataset, 
which is purged of various systematic errors and artifacts \cite{Hunt2024} 
and show that this ARG is highly congruent to an UShER phylogeny built from the same dataset. 
Compared to previously identified recombinants, we show that this ARG contains the vast majority of 
epidemiologically relevant signals of recombination, while adding significant nuance and texture. 
This ARG – which captures the major signals of evolution 
through phylogenetic relationships, point mutations, deletions, and recombination – 
is a resource of unprecedented detail and precision, 
and should provide a strong basis for future studies of SARS-CoV-2.

@@

%%% The results may be divided with subheadings. 
%%% In our view, good subheadings convey information about the findings, 
%%% so we encourage you to be specific. For example, "Factor X requires factor Y to function in process Z" 
%%% is better than "Analysis of factors X and Y by approach Q." 
%%% We recommend that you use similar language in your figure titles for clarity and structural harmony. 
%%% Please do not use numbered subheadings. For some formats, ``results and discussion" may be combined under a single heading.

%%% For initial submissions, figures and their titles and legends can be run-in with the text.
%%% For final submissions, the image files must be uploaded as separate files, one complete file per figure. The figure titles and legends should be placed toward the end of the main text, after the declaration of interests. Our typesetters will place the figures and their titles and legends appropriately based on your in-text citations.

%%% To cite/link to display items (figures and tables) and/or sections of the manuscript, 
%%% simply write, for example, ``(Figure 1)" or ``see discussion." 
%%% Our team will do the rest. Please note that we do not number sections or subsections.

\subsection*{Brief description of sc2ts}

Sc2ts reconstructs an ARG by incrementally adding batches of sequences 
as they become available (Figure 1A; for details, see STAR Methods). 
First, sc2ts infers likely “copying paths” connecting each sample in the batch to node(s) in the current ARG 
(allowing for recombination) using the Li and Stephens (LS) model \cite{Li2003}, 
which is a Hidden Markov Model (HMM) widely used in human genomics 
to approximate the effects of mutation and recombination on genetic inheritance. 
As recombination is relatively rare in viruses, 
we would expect most samples to copy from a single parent in the ARG. 
Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path 
between the parents resulting in different genomic segments inherited from the different parents. 
Next, for clusters within the batch that are inferred to copy from the same ARG node, 
sc2ts uses standard phylogenetic techniques to resolve relationships within the cluster. 
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, 
and applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.

Sc2ts uses collection dates to build a time-resolved ARG. 
When examining preliminary ARGs, we found samples that introduced large numbers of mutations (and reversions), 
likely due to errors in their collection dates 
(for example, a large batch of samples dated 31 December, 2020 contains samples from Omicron BA.1, 
which is estimated to have emerged in November, 2021; REF). 
To preclude adding such “time travelling” samples, we filtered samples by “HMM cost”. 
We define the “HMM cost” of a sample to be the total penalty cost 
for switches and mismatches (mutations) in its copying path. 
We parameterise the LS model such that each mismatch has a penalty cost of one 
and a switch has a penalty cost of \textit{k} (here, set to 4), 
which determines how much mutation is favored over recombination when accounting for a sample (Figure 1B). 
If the HMM cost of a sample exceeds a threshold (here, set to 7), 
sc2ts does not immediately add the sample to the ARG.

This approach resulted in preliminary ARGs that had noticeably fewer time travellers, 
but we found that it also excluded the major “saltation” events that characterize many VOCs. 
These saltation lineages are characterised by a high number of accumulated mutations, 
and are hypothesised to emerge from chronic infections 
during which the virus is exposed to individual-specific immune selective pressures \cite{Markov2023}. 
Though uncommon, such lineages drove large epidemic waves during the pandemic (notably Alpha, Delta, and Omicron). 
To distinguish time travellers with erroneous dates from true saltation lineages, 
we reevaluate the evidence for each lineage with a high HMM cost as additional samples are added. 
If on subsequent days these lineages form a sufficiently large group of samples, 
sc2ts adds the group back to the ARG. 
This “retrospective sample grouping” successfully captured the emergence of most saltation lineages (Figure SX). 
The emergence of the Delta, BA.1 and BA.2 lineages required some manual intervention, 
in which specific "seed" samples were identified for unconditional inclusion 
(for details of how seed samples were chosen, see STAR Methods). 
First arising in geographic regions with low sampling density, 
the initial origins of these VOC are challenging to infer (REF).

\noindent\includegraphics[width=0.85\linewidth]{methodology.png}

\subsection*{Figure 1: Schematic overview of sc2ts.} 
(A) Procedure to build an ARG by incrementally updating it with samples sorted by collection dates. 
Sample 0 is used to initialise the ARG as the root. 
Next, Sample 1 is attached, mapping mutations onto the new branch to explain it. 
When Sample 4 is detected as a recombinant, 
a node representing the recombination event is inserted with two edges connecting to its parents, 
with Sample 4 itself attached to this node as its child. 
The resulting ARG contains two local trees, one on each side of the inferred breakpoint. 
(B) Copying paths for Sample 4 inferred under the LS model 
while allowing recombination (\textit{k} = 4) or disallowing it (\textit{k} = 1000).

\subsection*{A comprehensive representation of SARS-CoV-2 evolution}

Using sc2ts, we built an ARG of ~2.48 million-samples spanning the COVID-19 pandemic. 
Inference took 21 days on a server equipped with dual EPYC (details) 
and 500 GB of RAM (~5 CPU years; Figure SX). 
The ARG contains 2,689,054 nodes representing the genealogical relationships 
among 2,482,157 sampled genomes in the Viridian v04 dataset, 
which were collected from January 01, 2020 to February 20, 2023 \cite{Hunt2024}. 
Of the nodes, 929 represent recombination events, 
of which 386 have a minimum of four nucleotides supporting the non-focal parent on both sides of the breakpoint 
(we refer to these as “robust” recombination events, Figure S1; see Document S1), 
reflecting the overwhelming tree-like structure of SARS-CoV-2 evolutionary history.

The ARG represents the full evolutionary history of SARS-CoV-2 
and is in close agreement with the state-of-the-art UShER phylogeny built using the Viridian data \cite{Hunt2024}. 
First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024} 
by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods). 
Tanglegrams reveal that the ARG captures known phylogenetic relationships both within sublineages 
and at the level of major lineages (including VOCs), 
with good agreement between the phylogenetic bipartitions in the ARG and the UShER phylogeny (Figure S2). 
Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian data 
by pruning both to their intersection of 2,475,418 samples and 27,431 sites. 
Remarkably, given the differences in inference methods and data processing pipelines, 
they are almost identically parsimonious, 
with sc2ts requiring 4,208 (0.23\%) fewer mutations than UShER (1,843,132 for UShER vs 1,838,924 for sc2ts). 
We also compared the mutational spectra of major VOCs in the ARG 
against the mutational spectra computed from the UShER phylogeny (STAR Methods), 
finding a close match to previous reports \cite{Bloom2023} (Figure S3).

Input genome alignments are masked in sc2ts 
so that deletions and ambiguous bases are treated as missing data by the HMM, 
and once the samples are incorporated into the ARG, 
these bases are imputed to carry the values inherited by the node in question. 
Deletions are an important part of SARS-CoV-2 evolution [cites], however, 
and to incorporate them into the ARG we identified a set of [XXX] sites most frequently involved in deletions [cite source?] 
and post-hoc mapped alignment data for these sites to the ARG using parsimony. 
Although sites are treated independently in this approach, 
the approximation successfully identified the most frequent deletions, 
such as the recurrent 21991-21993 (see Document S1 for more details). 
The node representing the origin of Alpha, with 290,407 descendants, 
is correctly associated with key lineage-defining deletions: 
11288-11296 (ORF1A del3675-3677), 
21765-21770 (Spike del69-70), 
21991-21993 (Spike del144), and 
28271 (non-coding). 
The Spike 69-70 deletion, which causes S-Gene Target Failure (REF), 
also emerges twice in the origin of Omicron - 
in the node leading to the BA.1 variant (342,099 descendants), 
and the node leading to the BA.4 variant (189,746 descendants). 
To speed up inference and to avoid issues caused by sites enriched for errors [citation?], 
sc2ts masks out a set of 100 sites identified as the most highly homoplasic 
by running on a subset of the data (STAR Methods). 
Of these sites, 9 are in the set of [1XX] “problematic sites” masked by UShER [citation?]. 
In addition to remapping the data for sites identified as being involved in deletions, 
we added mutations for XX of these highly recurrent sites 
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony. 
Each node in the ARG was then assigned to one of 2026 Pango lineages using pangolin (STAR Methods). 
All but 8293 sampled sequences (0.33\%) agreed with the source sequences in terms of Pango lineage assignment 
(comparable to the 0.27\% disagreement between pangolin-data versions 1.21 and 1.29 in the source Viridian metadata). 
A majority of these lineages (1469 of 2058) have monophyletic origins within the ARG, 
increasing to 1779 (86\%) when we allow for multiple sibling origination nodes (Table SX; Document S1).

\subsection*{Epidemiologically relevant recombinant lineages}

As new data became available during the pandemic, 
volunteers continually detected recombinants which may develop concerning features, 
particularly increased transmissibility, disease severity, 
and ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2022}). 
The process of gathering evidence for and characterising such recombinants is manual, 
involving visual inspection of phylogenies for unusual branches 
caused by forcibly placing recombinant sequences onto them, 
typically followed by visual analysis of genome sequence alignments 
to identify parental lineages and breakpoint locations. 
This process has resulted in the designation of recombinant Pango lineages 
(whose aliases begin with an “X”), to which we refer as “Pango Xs”. 
Below, we summarise our analysis of the Pango Xs in the ARG 
(for extended analyses, see Documents S1 and S2).

We find strong evidence for the recombination events leading to 17 Pango Xs in the ARG (Table 1), 
which passed the detection-level sc2ts quality control filters (Figure S1). 
These events are supported by high numbers of averted mutations (Table 1) 
and high numbers of loci matching their suggested parents (see their copying patterns in Figure SX). 
Among these Pango Xs are XA, which is the first designated Pango X showing signs of community transmission \cite{Jackson2021}, 
and XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2022}. 
The recombinant origin of XA is well explained in the ARG by two parents and one mutation (Figure 2A; Table 1), 
with sc2ts correctly identifying its parental Pango lineages and breakpoint location.

Furthermore, some of the Pango Xs above originate from shared recombination events in the ARG (Table 1). 
An example of a set of “nested” closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP, 
which descend from a single recombination event rather than five independent events in the ARG (Figure 2B). 
These Pango Xs were designated around the same time. 
These results highlight that the ARG can help clarify the relationships between recombinants, 
allowing for more precise interpretation of evidence for recombination.

Sc2ts identified 12 Pango Xs in the ARG as non-recombinants (Table 1), 
on the basis of the HMM and available data. 
For all these Pango Xs, the number of recombination-informative sites 
fell below our threshold of detection (i.e. k = 4 in the HMM). 
The fact that most of these Pango Xs (except XAJ and XAN) differ from a single parent node in the ARG 
by at most three mutations (Table 1) strongly supports their origins as non-recombinants. 
For example, the closely related XN and XAU can be explained 
by two mutations and one mutation in the ARG, respectively (Figure 2C). 
%%% TODO: Say something about XAJ and XAN.
However, sc2ts did not detect several Pango Xs (XB, XP, and XAJ) as recombinants 
because the HMM treated characteristic deletions as missing data.

\noindent\includegraphics[width=0.85\linewidth]{pango_x_subgraphs.png}

\subsection*{Figure 2: Illustrative local subgraphs in the ARG.}
Subgraphs around the XA-associated recombination node (depicted as an encircled R) (A); 
the recombination node associated with a group of “nested” Pango Xs (B); 
and the MRCA of the putative non-recombinants XN and XAU (C). 
For more detailed visualisations of these subgraphs, see Document S2.

\subsection*{Table 1. Recombination events associated with Pango Xs.}

\subsection*{Comparison with other recombination detection methods}

We compared sc2ts against three state-of-the-art (non-phylogenetic) methods 
that detect inter-lineage recombinants by scanning sequences for mutations associated with different Pango lineages: 
(1) RecombinHunt \cite{Alfonsi2024}, (2) CovRecomb \cite{Li2024}, and (3) rebar (https://github.com/phac-nml/rebar).

To evaluate the concordance of the methods in identifying the parental Pango lineages 
and locating the breakpoints of the recombinants, we focus on the Pango Xs, using the data 
from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison. 
For RecombinHunt, we took the detection results for the Pango Xs presented in the original study; 
for CovRecomb, we summarised the results from the supplementary data of the original study (STAR Methods); 
and for rebar, we obtained results by using it to classify the Pango X-associated recombination events in the ARG (STAR Methods). 
For each method, we determined whether the inferred parental lineages 
and breakpoint intervals are concordant with those identified by the community (STAR Methods). 
We find that the methods performed well overall, 
with no significant difference in concordance among them (Table 1; p-value = XX, chi-square test), 
indicating that sc2ts can characterise recombinants as well as the state-of-the-art recombination detection methods.

We also compared the sensitivity of sc2ts and rebar. 
Of all the “robust” recombination events detected by sc2ts, 
167 (43\%) were also detected as recombinants by rebar (STAR Methods). 
We hypothesise that this disagreement arose from differences in sensitivity between these methods. 
Consistent with this, we find that rebar needs a higher level of divergence 
between parents to correctly classify a sequence as recombinant, 
as shown by a shorter time to their MRCA (Figure 3A) 
and lower distance between their Pango identity (Figure 3B).

\subsection*{Table 2. Concordance among methods in detecting Pango Xs.}

\begin{tabular}{|l | l | l | l|} 
 \hline
 \textbf{Method} & \textbf{Concordant in parental lineages (\%)} & \textbf{Concordant in breakpoint interval (\%)} & \textbf{Distance (bp)} \\ [1ex] 
 \hline
 Sc2ts & XX & XX & XX \\ [1ex] 
 \hline
 RecombinHunt-GISAID & XX & XX & XX \\ [1ex] 
 \hline
 RecombinHunt-Nextstrain & XX & XX & XX \\ [1ex] 
 \hline
 CovRecomb & XX & XX & XX \\ [1ex] 
 \hline
 Rebar & XX & XX & XX \\ [1ex] 
 \hline
\end{tabular}

\subsection*{Quantification of evidence for recombination}

%%% SHZ: We may not need this section at all, but I'm keeping it for now.
%%% We should probably have a condensed version of the first paragraph in the Pango X section.
%%% The second and third paragraphs should probably be removed entirely.

Sc2ts provides a method to facilitate the discovery of recombination events, 
by comparing how well a sample fits the ARG under a model allowing recombination 
versus a model disallowing recombination, 
as measured by the number of resulting mutations (Figure 1B). 
The more mutations are “averted” by allowing recombination, 
the stronger the evidence for recombination. 
We calculated the number of averted mutations for all the “robust” recombination events (STAR Methods), 
which include all the Pango X-associated recombination events (Table 1).

Our analysis of the number of averted mutations reveals a spectrum of plausibility 
across all the “robust” recombination events (Figure 3C). 
Examples of highly plausible recombinants are XA and XBB (for an extended analysis, see Document S1). 
The “copying pattern” of XA visually confirms that a high number of recombination-informative sites 
support the suggested parents (Figure 3D). 
XW is an example of a marginally well-supported recombinant. 
The initial evidence for proposing XW for designation is four nucleotides associated with the suggested BA.1 parent. 
The copying patterns of all the recombination events are shown in Document S3.

Our results suggest that Delta and Omicron BA.2 might have recombinant origins, 
but we interpret these results with caution, 
as these VOCs are undersampled around the time of their emergence (REF). 
Consistent with this, the presence of mutations and conflicting loci matching non-focal parents 
observed in the copying patterns of these VOCs (Figure 3D) suggests the existence of 
many genetically distinct, unsampled ancestors. 
Future work is needed to test whether these VOCs descended from recombination events, 
with improved sampling if still possible.

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\noindent\includegraphics[width=0.85\linewidth]{averted_mutations.png}

\subsection{Figure 3: Evidence for recombination events in the ARG.} 
The 386 “robust” recombination events detected by sc2ts are colored 
according to their classification by rebar (recombinant: purple, non-recombinant: gray). 
(A) Distribution of time to the MRCA of the sc2ts-identified parents. 
(B) Distribution of the Pango lineage distance between the sc2ts-identified parents. 
(C) Distribution of the number of mutations averted by recombination (bottom x-axis), 
defined as the difference in the number of mutations 
between a copying path allowing recombination (right y-axis) 
and an alternative copying path disallowing recombination (top x-axis). 
(D) Copying patterns of five select recombinants.

\subsection*{Recombination breakpoint intervals along the genome}

We explored the spatial distribution of the breakpoint locations of 
the “robust” recombination events along the genome, 
and quantified how precisely breakpoints were localized. 
For each event, we computed an interval estimate of its breakpoint location, 
using its size as a measure of precision (median, 2,207 bases; STAR Method).

We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome, 
particularly at the left boundary of the Spike gene (Figure 4A), 
consistent with the pattern of excess observed in previous studies \cite{Turakhia2022,Li2024}. 
While this might result from a higher recombination rate in this region, 
it may also be due to the higher levels of polymorphism in the Spike gene, 
allowing recombination endpoints to be estimated more precisely. 
Reassuringly, sites having unusually high mutation counts are absent in this region (Figure 4B), 
which would confound recombination detection.

Furthermore, we find an intrinsic negative relationship 
between breakpoint interval size and the amount of divergence between recombinant parents, 
which is measured by time to their MRCA (Figure 4D). 
When the two parental lineages have diverged for longer periods of time (the rightmost points), 
more information becomes available on either side of a breakpoint to pinpoint its location, 
resulting in shorter intervals. 
For instance, the breakpoint intervals tend to be the smallest for events 
involving recombination between the deeply divergent Omicron BA.1 and BA.2 sublineages 
(the red points in Figure 4D).

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\noindent\includegraphics[width=0.85\linewidth]{recombination_landscape.png}

\subsection*{Figure 4: Recombination landscape.} 
(A) Distribution of the breakpoint intervals of the 386 “robust” recombination events along the genome 
(annotated with gene labels on top). Each line represents an interval colored by its size. 
(B) Number of overlapping breakpoint intervals (orange; left y-axis) 
and number of mutations along the genome (blue; right y-axis). 
The coordinates of the ten sites with the highest mutation counts are labelled. 
(C) Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations. 
(D) Relationship between breakpoint interval size and time 
between the recombination event and the parental MRCA. 
The black curve shows the theoretical expected relationship between these quantities, 
estimated by minimizing the sum of squared residuals given a single parameter (the mutation rate) (STAR Methods).

\subsection*{Emergence of recombinants over time}

We examined the frequency of emergence of the recombinants over the time course of the COVID-19 pandemic (Figure 5A). 
We explored two hypotheses for the timing of recombination. 
First, we would expect recombination events to occur more often when case counts are high, 
which increases the chance of co-infection within an individual, allowing recombination. 
Second, we would expect recombination events to be detected more often when lineage diversity is high, 
as this increases the chance that two sufficiently distinct viruses co-infected an individual, 
increasing our power to detect recombination.

We used the COVID-19 Data Repository at Johns Hopkins University \cite{Dong2020} 
to relate the timing of recombination events with global case numbers. 
Case numbers (\textit{n}) were binned by week from January, 2020 to March, 2023 
and sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in Figure 5A). 
These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset. 
Alongside the case numbers, the black curve shows the number of recombinants inferred within each week along the ARG, 
focusing on the 386 “robust” events. 
Recombination events were inferred more often during the Delta and Omicron waves, when case counts were high. 
Overall, the number of recombinant events rose with the global number of cases 
(Figure 5B; linear regression: 1.2 + 2.9 10^{-7} x, p = 3.2  10^{-7}). 
Recombination events also rose with the diversity of lineages in each week 
(Figure 5C; linear regression: 1.0 + 5.3 x, p = 1.3 10^{-7}), 
measured as the chance that two randomly drawn viruses have different Scorpio designations 
(H, referred to as the “expected heterozygosity”).

We then asked which single measure best explains the number of recombination events inferred each week, 
considering \textit{n}, \textit{H}, \textit{n H}, \textit{n}^2, \textit{n}^2 \textit{H}. 
We included the squared number of cases each week thinking that it might better predict co-infections. 
The best single predictor of recombination events involved the product of cases and expected heterozygosity (\textit{n H}), 
with an adjusted R^2 of 25.3\%. 
This measure of case diversity (\textit{n H}) explained 62\% more of the variation in recombination events per week 
than the next best predictor (\textit{H}, adjusted R^2 of 15.7\%). 
While consistent with the hypotheses that recombination events occur more often when infection rates are common 
and are detected more often when viral diversity is high, we note several sources of uncertainty. 
Global case numbers were underreported, particularly during the Delta and Omicron waves \cite{Wang2025}. 
Furthermore, there may be substantial uncertainty in the timing of a recombination event, 
particularly if these occur in chronically infected individuals (as reported by \cite{Burel2022}) 
with substantial delays before onward cases are detected. 
Finally, the filters that we applied removed all but the most strongly supported recombination events, 
although similar conclusions were obtained when all 929 recombinants were considered (STAR Methods).

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\noindent\includegraphics[width=0.85\linewidth]{recombinants_over_time.png}

\subsection*{Figure 5: Recombination events across the pandemic.} 
(A) Timing of the recombination events (black curve, right y-axis) 
relative to the global case numbers (\cite{Dong2020}; left y-axis). 
Cases are color-coded by their Scorpio labels, 
with the key showing those groupings with ≥ 1,000 samples in the Viridian dataset. 
Recombinants more commonly occur (B) in weeks with high case numbers 
and (C) in weeks with extensive lineage variation, 
as estimated by high expected heterozygosity (C); 
solid lines show fits by linear regression, 
allowing for non-zero intercepts to account for unknown cases 
and heterozygosity not captured by the Scorpio labels.

\subsection*{Efficient inference and analysis of a pandemic-scale ARG}

The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale, 
which overwhelmed existing computational approaches, 
motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}. 
Although the computational tools used in phylogenetics and population genetics have historically been largely distinct, 
the functionality required largely overlaps. 
Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024} 
and the VCF Zarr data storage format \cite{Czech2025} 
to describe relationships among virus samples, 
building an ARG over time as data becomes available. 
By reusing high-quality and high-performance components of these toolkits, 
we could develop the sc2ts inference method itself 
and perform all the pandemic-scale analyses in this article with Python code alone and no compiled languages.

To illustrate the power of these libraries we provide some illustrative examples 
in which we compare the Usher and sc2ts intersections in tskit format described earlier. 
First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments. 
The total aligned dataset for 1,226,587 samples and 27,431 sites shared by the sc2ts and Usher tree sequences 
represents 31.34 GiB of nucleotide calls. 
Of the 80,574,518 missing data calls (Ns) in the alignments, 
sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%). 
Additionally, 413,303 calls made use of the IUPAC uncertainty codes. 
Of these sc2ts imputed only 77 (0.02\%) incorrectly (i.e., with a base that is not compatible with the ambiguity code). 
UShER imputed only 4 calls from this set incorrectly, 
reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations. 
This computation, performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees, 
required 30 lines of Python code and took 2m59s on an intel i7-9700 CPU, with a peak memory usage of about 4.5G 
(see \cite{Czech2025} for further benchmarks for processing the Viridian data in VCF Zarr format). 
Performing this analysis using matUtils would require exporting the full dataset to VCF, 
and converting the VCF to FASTA, and then comparing the alignments one-by-one to the source FASTA alignments. 
In contrast with matUtils and most phylogenetic software, 
which provide a command line interface and suite of predefined operations that typically output text files, 
tskit is a Python API (with C and Rust interfaces also supported) which has many advantages \cite{Baumdicker2022}. 
As an illustration of the open-ended and efficient nature of the analyses this enables, 
we computed the number of polytomies (nodes that have > 2 children) 
and the maximum number of children per node in the Usher and sc2ts outputs. 
In the Usher tree it took 116ms and 4 lines of code to count polytomies (171,148) 
and compute the maximum number of children (5918). 
In the sc2ts ARG, it took 385ms and 6 lines of code to find the maximum number of polytomies (137,001) 
and maximum number of children (7708) across all 348 local trees.

@@

%%% The discussion should explain the significance of the results 
%%% and place them into a broader context. Subheadings are permitted.

We present sc2ts, a real-time ARG inference method, 
and applied it to build a ~2.48 million-sample pandemic-scale ARG of SARS-CoV-2. 
By identifying specific samples as parents, 
sc2ts enables more precise interpretation of recombinants than state-of-the-art recombination detection methods, 
which find recombinants by matching a given sample to pre-defined population-averaged mutational profiles of Pango lineages. 
Sc2ts treats recombination as an integral part of SARS-CoV-2 evolution, in contrast to RIPPLES \cite{Turakhia2022} 
which detects recombinants post hoc by searching for unusual branches in a given phylogeny, 
inevitably resulting in an incomprehensive representation of SARS-CoV-2 evolution. 
Combined with new methods and tools to evaluate evidence for recombination, 
sc2ts is a significant technological advance over current methods for the discovery 
and analysis of recombinants and efficient inference of SARS-CoV-2 genealogy.

The ARG itself is a useful resource to support future studies of SARS-CoV-2. 
Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution, 
which are shown in different studies using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}. 
The ARG displays remarkable phylogenetic congruence with a state-of-the-art phylogeny built using UShER. 
Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes, 
allowing for convenient and efficient reuse of this valuable dataset 
for a variety of evolutionary and genetic analyses beyond those presented here. 
We hope that the ARG will spur on new research areas, in which recombination is naturally and fully accounted for, 
and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022} 
has made on our understanding of human genetic ancestry and demographic past.

We contribute several novel methods and concepts which may find applications beyond sc2ts. 
First, we devise a simplified LS model, which extends the classic concept of parsimony. 
In sc2ts, this model allows for accelerated HMM likelihood calculations, 
enabling reconstruction of the pandemic-scale ARG presented here. 
Existing HMMs used in statistical genetics may be modified in a similar fashion for significant speedup. 
Second, we propose the concept of mutations averted by recombination, 
which leads to an interpretable quantity to weigh 
the support for the hypothesis that a given sample is a recombinant 
against the hypothesis that it is a non-recombinant, 
considering the sampled genetic diversity in an ARG and a specific HMM. 
Third, we propose the concept of HMM cost to identify poorly explained samples 
which are either too divergent or have a high number of mutations inconsistent with their collection dates. 
Here we use HMM cost to reject time travelling samples, 
which no existing phylogenetic method detects and handles during inference 
(e.g., in the Viridian UShER phylogeny, parent and child nodes do not necessarily follow time order \cite{Hunt2024}), 
resulting in a clean dataset freed of samples with erroneous collection dates that may affect downstream analyses. 
Future work may be estimating dates for the suspected time travelling samples using the ARG provided here 
and the sc2ts sample matching engine.

Recombination is an important evolutionary force in viruses \cite{Pérez-Losada2015}, 
but it has been understudied partly owing to a lack of suitable data models, methods, and tools \cite{Neches2020}. 
While the methods and tools developed here are tailored for SARS-CoV-2, 
they can be extended and adapted for other viruses.
A major advantage of the sc2ts methods and tools is 
that they are built on top of the mature software libraries of tskit and VCF-Zarr,
which enable pandemic-scale analyses that can be performed on typical laptops. 
Both tskit and VCF-Zarr are developed in Python and leverage the popular PyData ecosystem, 
making them highly adaptable for a broad range of applications. 
Together with the pandemic-scale ARG, we hope that sc2ts will be highly accessible to many researchers, 
enabling and accelerating studies of SARS-CoV-2 for years to come.

@@

%%% A "limitations" or "limitations of the study" subsection in the discussion is encouraged 
%%% and may be required for some journals and some article formats.
